---
title: "In-class Exercise 5"
editor: visual
---

## Overview

This is in-class exercise 5.

## Getting Started

```{r}
pacman::p_load(tidyverse, funModeling, blorr, corrplot, ggpubr, sf, spdep, GWmodel, tmap, skimr, caret)
```

## Geospatial Data Wrangling

### Importing Geospatial Data

```{r}
osun <- read_rds("rds/Osun.rds")
```

```{r}
osun_wp_sf <- read_rds("rds/Osun_wp_sf.rds")
```

```{r}
osun_wp_sf %>% freq(input = 'status')
```

```{r}
tmap_mode("view")
tm_shape(osun) + tm_polygons(alpha = 0.4) + tm_shape(osun_wp_sf) + tm_dots(col = "status", alpha = 0.6) + tm_view(set.zoom.limits = c(9,12))
```

Summarize and give a good view of the data with just one line of code.

```{r}
osun_wp_sf %>% skim()
```

Checking for missing values from multiple variables and change the data type of usage_capacity from numeric to factor as it will be treated as continuous data type during logistic regression when it should be categorical

```{r}
osun_wp_sf_clean <- osun_wp_sf %>% 
  filter_at(vars(status,
                 distance_to_primary_road,
                 distance_to_secondary_road,
                 distance_to_tertiary_road,
                 distance_to_city,
                 distance_to_town,
                 water_point_population,
                 local_population_1km,
                 usage_capacity,
                 is_urban,
                 water_source_clean), all_vars(!is.na(.))) %>% mutate(usage_capacity = as.factor(usage_capacity))
```

Drop the geometry column of the simple feature data frame.

```{r}
osun_wp <- osun_wp_sf_clean %>% select(c(7, 35:39, 42:43, 46:47, 57)) %>% st_set_geometry(NULL)
```

Checking for highly correlated variables.

```{r}
cluster_vars.cor = cor(osun_wp[,2:8])
corrplot.mixed(cluster_vars.cor,
         lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

```{r}
model <- glm(status ~ distance_to_primary_road +
                 distance_to_secondary_road +
                 distance_to_tertiary_road +
                 distance_to_city +
                 distance_to_town +
                 water_point_population +
                 local_population_1km +
                 usage_capacity +
                 is_urban +
                 water_source_clean, data = osun_wp_sf_clean, family = binomial(link = 'logit'))
```

Showing the result of the model

```{r}
model
```

```{r}
blr_regress(model)
```

```{r}
blr_confusion_matrix(model, cutoff = 0.5)
```

Sensitivity = TP/(TP + FN)

Specificity = TN/(TN + FP)

Accuracy = (TP + TN)/(TP + TN + FP + FN)
