---
title: "In-class Exercise 4"
editor: visual
---

## Overview

This is in class exercise 4

## Getting Started

olsrr package:

-   Build regression model

-   Allow calibration of regression models using different methods

-   Give a more tidy report to see the overall performance of the model

corrplot:

-   multi-collinearity analysis

```{r}
pacman::p_load(olsrr, corrplot, ggpubr, sf, spdep, GWmodel, tmap, tidyverse, gtsummary)
```

## Geospatial Data Wrangling

### Importing geospatial data

```{r}
mpsz = st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

### Updating CRS information

```{r}
mpsz_svy21 <- st_transform(mpsz, 3414)
```

## Aspatial Data Wrangling

### Importing the aspatial data

```{r}
condo_resale = read_csv("data/aspatial/Condo_resale_2015.csv")
```

### Converting aspatial data frame into a sf object

Simple feature data frame where the longitude and latitude is combined to a coordinate field

```{r}
condo_resale.sf <- st_as_sf(condo_resale,
                            coords = c("LONGITUDE", "LATITUDE"),
                            crs=4326) %>%
  st_transform(crs=3414)
```

### Drawing Statistical Point Map

```{r}
tmap_mode("view")
```

```{r}
tm_shape(mpsz_svy21) + tmap_options(check.and.fix = TRUE) + tm_polygons(alpha = 0.1) +
tm_shape(condo_resale.sf) +  
  tm_dots(col = "SELLING_PRICE",
          alpha = 0.6,
          style="quantile") +
  tm_view(set.zoom.limits = c(11,14))
```

## Hedonic Pricing Modelling in R

### Simple Linear Regression Method

```{r}
condo.slr <- lm(formula=SELLING_PRICE ~ AREA_SQM, data = condo_resale.sf)
```

```{r}
summary(condo.slr)
```

Look at the adjusted R square to see the fit of the line to the data.

### Multiple Linear Regression Method

Cannot use simple feature for corrplot. Otherwise will have error as the object it accept is a normal data frame.

```{r}
corrplot(cor(condo_resale[, 5:23]), diag = FALSE, order = "AOE",
         tl.pos = "td", tl.cex = 0.5, method = "number", type = "upper")
```

### Building a hedonic pricing model using multiple linear regression method

```{r}
condo.mlr <- lm(formula = SELLING_PRICE ~ AREA_SQM + AGE    + 
                  PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE +
                  PROX_URA_GROWTH_AREA + PROX_HAWKER_MARKET + PROX_KINDERGARTEN + 
                  PROX_MRT  + PROX_PARK + PROX_PRIMARY_SCH + 
                  PROX_TOP_PRIMARY_SCH + PROX_SHOPPING_MALL + PROX_SUPERMARKET + 
                  PROX_BUS_STOP + NO_Of_UNITS + FAMILY_FRIENDLY + FREEHOLD, 
                data=condo_resale.sf)
summary(condo.mlr)
```
